Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-07-30 manuscript/manuscript.tex
edits

A ubiquitous feature of repeated measures studies is loss of data due to intermittent missingness and attrition. In the Bayesian setting, the standard approach to dealing with missing data is to perform multiple imputation, whereby $m$ imputed data sets are generated from a specified imputation model. After $m$ complete data sets are obtained, parameter estimates are combined across each data set to produce a final set of parameter estimates (Gelman \textit{et al.} 2013). This approach is not only computationally burdensome, requiring storage and analysis of an $m \times n_{rows} \times n_{cols}$ data array in addition to multiplication of total model run time by a factor of $m$, but it has been shown to produce unreliable inferences (Zhou and Reiter, 2010). We instead include an ``online" imputation imputation step in our Gibbs sampling procedure, whereby missing outcomes are updated at each iteration. This approach greatly increases the number of opportunities for exploration of the missing data parameter space and avoids the multiplication of total run time and number of parameters.

@@

We can extend model (\ref{eq:msndens}) to the regression setting by modeling $\boldsymbol\zeta_{ki}$ as a function of covariates. Here we adopt a convenient stochastic representation of the MSN density (Azzalini and Dalla Valle, 1996):
\begin{equation}
\mathbf{y}_{i}|(z_i=k,t_i) = \mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k + \boldsymbol\epsilon_{ki}, \label{eq:msnreg}
\end{equation}
where 
$\mathbf{X}_i$ is a $J \times Jp$ design matrix that includes potential time-varying covariates (e.g., indicators denoting quarterly visits); $\boldsymbol{\beta}_k=(\beta_{k11},\ldots,\beta_{k1p},\ldots,\beta_{kJ1},\ldots,\beta_{kJp})^T$ is a $Jp\times 1$ vector of cluster- and outcome-specific regression coefficients; $t_i\sim \text{N}_{[0,\infty)}(0,1)$ is a subject-specific standard normal random variable truncated below by zero; $\boldsymbol\psi_k=(\psi_{k1},\ldots,\psi_{kJ})^T$ is a $J \times 1$ vector of cluster-specific skewness parameters; and $\boldsymbol{\epsilon}_{ki} \sim \text{N}_J(\boldsymbol0,\boldsymbol\Sigma_k)$ is a $J\times 1$ vector of correlated error terms. Thus, conditional on $t_i$ and $z_i=k$, $\boldsymbol{y}_i$ is distributed as $\text{N}_J(\mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k, \boldsymbol{\Sigma}_k)$. Marginally (after integrating over $t_i$), $\mathbf{y}_i|(z_i = k)$ is distributed $\text{MSN}_J(\boldsymbol\zeta_{ki}, \boldsymbol\alpha_k, \boldsymbol\Omega_k)$, where through back-transformation 
\begin{eqnarray}
\boldsymbol\zeta_{ki} &=& \mathbf{X}_i\boldsymbol\beta_k, \nonumber \\
\boldsymbol\alpha_k &=& \frac{1}{\sqrt{1 - \boldsymbol\psi_k^T 
\boldsymbol\Omega^{-1}_k\boldsymbol\psi_k}} \boldsymbol\omega_k \boldsymbol\Omega^{-1}_k\boldsymbol\psi_k,~~\text{and} \label{eqn:back_transform}\\
\boldsymbol\Omega_k &=& \boldsymbol\Sigma_k  + \boldsymbol\psi_k \boldsymbol\psi_k^T \nonumber,
\end{eqnarray}
where $\boldsymbol\omega_k = \text{Diag}(\boldsymbol\Omega_k)^{1/2}$ is the $J \times J$ diagonal matrix containing the square root of the diagonal entries of $\boldsymbol\Omega_k$. Additional details can be found in Fr\"{u}wirth-Schnatter and Pyne (2010).

Of note, the MSN density can be expressed more compactly in terms of the matrix skew normal (MatSN) density (Chen and Gupta 2005). Let $\mathbf{Y}_k$ be an ${n_k \times J}$ response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n \mathds{1}_{(z_i = k)}$ is the number of observations in cluster $k$. From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_k$ is distributed as
\begin{eqnarray}
\mathbf{Y}_{k} &\sim& \text{MatSN}_{n_k \times J}(\mathbf{M}_k,\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Omega_k)\label{eq:matsn}\\
%\text{vec}(\mathbf{Y}_k) =  (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})^T\\
\text{vec}(\mathbf{M}_k) &=& (\boldsymbol\zeta_{k1}^T,...,\boldsymbol\zeta_{kn_k}^T)^T, \nonumber
\end{eqnarray}
where $\mathbf{M}_k$ is an $n_k \times J$ location matrix with rows $\boldsymbol\zeta_{ki} = \mathbf{X}_i \boldsymbol\beta_k$ as in equation (\ref{eq:msnreg}), $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha_{kJ})^T$, $\mathbf{I}_{n_k}$ is the $n_k\times n_k$ identity matrix, and $\boldsymbol\Omega_k$ is the $J\times J$ scale matrix defined above in equation (\ref{eq:msndens}). From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_{k}$, conditional on the $n_k\times 1$ vector of random effects $\mathbf{t}_{k} = (t_1,...,t_{n_k})^T$, is jointly distributed in matrix form as
\begin{equation}
\mathbf{Y}_k | \mathbf{t}_k \sim \text{MatNorm}_{n_k \times J}(\mathbf{M}^*_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k),
\end{equation}
where $\text{MatNorm}_{n_k \times J}(\cdot)$ denotes a $n_k\times J$ matrix normal density, $\mathbf{M}^*_k$ is an $n_k \times J$ matrix such that $\text{vec}(\mathbf{M}^*_k) = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$ is an $n_kJ\times 1$ mean vector, $\mathbf{X}_k$ is an $n_kJ \times Jp$ design matrix, $\boldsymbol\beta_k$ is the $Jp\times 1$ vector of regression coefficients defined in equation (\ref{eq:msnreg}), and $\boldsymbol\Sigma_k$ is the $J\times J$ conditional covariance of $\boldsymbol\epsilon_{ik}$ given in equation (\ref{eq:msnreg}). As described in Section 3.6, the matrix representation of the MSN distribution admits convenient conjugate prior distributions for the regression parameters and scale matrices, which in turn leads to efficient Gibbs sampling for posterior inference.

\subsection{Multinomial Regression on Cluster Indicators}
\label{s:multinom}
To accommodate heterogeneity in the cluster-membership probabilities, we model $\pi_{ki}$ as a function of covariates using a multinomial logit model 
\begin{eqnarray}
\label{eq:pik}
\pi_{ki} = \Pr(z_i = k|\mathbf{w}_i) = \frac{\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k}}{\sum_{h = 1}^K \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}},~ k=1,\ldots,K,
\end{eqnarray}
where $\mathbf{w}_i$ is an $r\times 1$ vector of subject-level covariates, $\boldsymbol\delta_k$ is a $r\times 1$ vector of regression parameters associated with membership in cluster $k$. For identifiability purposes, we fix the reference category $k = K$ and set $\boldsymbol\delta_K = \mathbf{0}$. Under model (\ref{eq:pik}), $z_i|\boldsymbol\pi_i \sim \text{Multinomial}(1,\boldsymbol\pi_i)$, where $\boldsymbol\pi_i = (\pi_{1i},...,\pi_{Ki})$. As detailed below in Section \ref{s:postinf}, each cluster label $z_i$ is updated during MCMC estimation from its multinomial full conditional distribution and used in the remaining MCMC steps as the working cluster assignment for subject $i$. By allowing the cluster probabilities to vary across subjects, model (\ref{eq:mixture}) can be viewed as a \textit{mixture of experts} model, in which $\pi_{ki}$ acts as a \textit{gating function} controlling the prior probability of membership in cluster $k$, and $f(\mathbf{y}_i|\boldsymbol\theta_k)$ in equation (\ref{eq:mixture}) is the ``expert'' providing information on the within-cluster distribution of $\mathbf{y}_i$ (Bishop 2006).

@@

\begin{equation}
\frac{(\text{e}^{\eta})^a}{(1 + \text{e}^{\eta})^b} = 2^{-b}\text{e}^{\kappa \eta} \int_{0}^{\infty} \text{e}^{-{w}\eta^2/2}p({w}|b,c = 0)d{w}, \label{eq:pg2}
\end{equation}
where $\kappa = a - b/2$ and $p(w|b,c = 0)$ denotes a $\text{PG}(b,0)$ density. Polson {et al.} further show that the conditional distribution $p(w|b,c)$ results from an ``exponential tilting'' of the $\text{PG}(b,0)$ density, thus

@@

Polson et al. use these results to show that, for the logistic model, the Bernoulli likelihood can be written as a scale-mixture of normal densities with Polya-Gamma precision terms, $w$, resulting in a normal full conditional distribution for the logistic regression parameters. To extend the P\'olya--Gamma data augmentation approach to the multinomial setting, we first introduce the binary indicators $U_{ki}$, such that $U_{ki} = \mathds{1}_{(z_i = k)}$ -- i.e., $U_{ki}$ is an indicator variable taking the value $1$ if subject $i$ belongs to cluster $k$, and $0$ otherwise. The full conditional distribution of $\boldsymbol\delta_k$ is then given as  $$p(\boldsymbol\delta_k|\mathbf{U}_k,\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ki}^{U_{ki}}(1-\pi_{ki})^{1-U_{ki}},$$

@@

$$\pi_{ki} = P(U_{ki} = 1) = \frac{\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}}}{1 + \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}}} = \frac{\text{e}^{\eta_{ki}}}{1 + \text{e}^{\eta_{ki}}},$$

@@

	p(\boldsymbol\delta_k|\mathbf{z},\mathbf{W}, \mathbf{O}_k) \propto p(\boldsymbol\delta_k) \text{exp} \left [ - \frac{1}{2} (\mathbf{U}^*_k - \mathbf{W}\boldsymbol\delta_k)^T \mathbf{O}_k (\mathbf{U}^*_k - \mathbf{W}\boldsymbol\delta_k)\right].

@@

To accommodate intermittent missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' -- that is, as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness (i.e., observations missing not at random). Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate normal distribution given $(z_i,t_i,\mathbf{y}^{obs}_i)$:

@@

\subsubsection{Prior Specification}
We adopt a fully Bayesian inferential approach and assign prior distributions to all model parameters. Conveniently, all parameters admit conditionally conjugate priors, which greatly improves posterior computation via a data-augmented Gibbs sampler. To make use of the matrix normal representation introduced previously in Section 3, we define the matrix of regression parameters $\mathbb{B}^*_k$, where $\text{vec}(\mathbf{B}^*_k) = \boldsymbol\beta^*_k = (\boldsymbol\beta_k^T,\boldsymbol\psi_k^T)^T$. We assign $\mathbf{B}^*_k|\boldsymbol\Sigma_k$ a $\text{MatNorm}(\mathbf{B}^*_{0k},\mathbf{I}_{p+1},\boldsymbol\Sigma_k)$ prior, where $\mathbf{B}^*_{0k}$ is a matrix of location parameters such that $\text{vec}(\mathbf{B}^*_{0k})$ is a vector of prior location parameters for the components of $\boldsymbol\beta^*_k$, $\mathbf{I}_{p+1}$ is the $(p+1)$-dimensional identity matrix, and $\boldsymbol\Sigma_k$ is the covariance matrix defined in equation (\ref{eq:msndens}), for which we specify an $\text{IW}(\mathbf{V}_{0k},\nu_{0k})$ prior. This leads to a matrix-normal-inverse-Wishart joint prior for $\mathbb{B}^*_k$ and $\boldsymbol\Sigma_k$, which is the conjugate joint prior for the regression parameters in the matrix normal model given in equation (\ref{eq:matsn}). This conjugate prior specification induces convenient closed-form full conditional distributions that can be easily updated within our proposed Gibbs sampler.

@@

\paragraph{\textbf{Step 2:} Update of MSN Regression Parameters} We begin the update of MSN regression parameters by first updating $t_i$, the truncated normal random effect used in the stochastic representation given in equation (\ref{eq:msnreg}). For cluster $k$, we compute $A_k = (1 + \boldsymbol\psi_k^T \boldsymbol\Sigma^{-1}_k\boldsymbol
\psi_k)^{-1}$ using current values of $\boldsymbol\psi_k$ and $\boldsymbol\Sigma_k$. Next, for $i = 1,...,n_k$, we compute $a_i = A_k \boldsymbol\psi_k^T \boldsymbol\Sigma^{-1}_k(\mathbf{y}_i - \boldsymbol\zeta_{ki})$, where $\boldsymbol\zeta_{ki} = \mathbf{X}_i \boldsymbol\beta_k$. Finally, we update $t_i$ from $\text{N}_{[0,\infty)}(a_i, \mathbf{A}_k)$.  Repeat these updates for $k = 1,...,K$.

The remaining MSN regression parameters are updated from their full conditionals as follows. First, we form the $n_k \times (p + 1)$ matrix $\mathbf{X}^*_k$ by column-binding $\mathbf{X}_k$ and $\mathbf{t}_k$. For each cluster $k$ ($k = 1,...,K$), we update $\boldsymbol\Sigma_k$ from an $\text{IW}(\nu_k,\mathbf{V}_k)$ density, where $\nu_k = \nu_{0k} + n_k$ and\begin{eqnarray}
	\mathbf{V}_k &=&  (\mathbf{Y}_k-\mathbf{X}^*_k \mathbf{B}^*_k)^T (\mathbf{Y}_k - \mathbf{X}^*_k \mathbf{B}^*_k) + (\mathbf{B}_k^* - \mathbf{B}^*_{0k})^T (\mathbf{B}^*_k - \mathbf{B}^*_{0k}) + \mathbf{V}_{0k}. \nonumber
\end{eqnarray} 
We then make use of the matrix normal representation introduced in Section 3.2 to draw $\mathbf{B}^*_k$ from a $\text{MatNorm}_{p+1,J}(\mathbb{B}_k^*,\mathbf{L}_k^{*},\boldsymbol\Sigma_k)$ density, where
\begin{eqnarray}
	\mathbb{B}^*_k &=& \mathbf{L}_k^{*} (\mathbf{X}^{*T}_k \mathbf{Y}_k + \mathbf{I}_{p+1} \mathbf{B}^*_{0k}) \nonumber \\
	\mathbf{L}^*_k &=& (\mathbf{X}^{*T}_k \mathbf{X}^*_k + \mathbf{I}_{p+1})^{-1}. \nonumber
\end{eqnarray}
For efficient sampling of the matrix normal distribution, we use the \texttt{R} package \texttt{matrixsampling} (Laurent 2018).

>>>>

A ubiquitous feature of repeated measures studies is loss of data due to intermittent missingness and attrition. In the Bayesian setting, the standard approach to dealing with missing data is to perform multiple imputation, whereby $M$ imputed data sets are generated from a specified imputation model. After $M$ complete data sets are obtained, parameter estimates are combined across each data set to produce a final set of parameter estimates (Gelman \textit{et al.} 2013). This approach is not only computationally burdensome, requiring storage and analysis of an $M \times n_{rows} \times n_{cols}$ data array in addition to multiplication of total model run time by a factor of $M$, but it has been shown to produce unreliable inferences (Zhou and Reiter, 2010). We instead include an ``online" imputation imputation step in our Gibbs sampling procedure, whereby missing outcomes are updated at each iteration. This approach greatly increases the number of opportunities for exploration of the missing data parameter space and avoids the multiplication of total run time and number of parameters.

@@

We can extend model (\ref{eq:msndens}) to the regression setting by modeling $\boldsymbol\zeta_{ki}$ as a function of covariates. Here we adopt a convenient stochastic representation of the MSN density (Azzalini and Dalla Valle, 1996):
\begin{equation}
\mathbf{y}_{i}|(z_i=k,t_i) = \mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k + \boldsymbol\epsilon_{ki}, \label{eq:msnreg}
\end{equation}
where 
$\mathbf{X}_i$ is a $J \times Jp$ design matrix that includes potential time-varying covariates (e.g., indicators denoting quarterly visits); $\boldsymbol{\beta}_k=(\beta_{k11},\ldots,\beta_{k1p},\ldots,\beta_{kJ1},\ldots,\beta_{kJp})^T$ is a $Jp\times 1$ vector of cluster- and outcome-specific regression coefficients; $t_i\sim \text{N}_{[0,\infty)}(0,1)$ is a subject-specific standard normal random variable truncated below by zero; $\boldsymbol\psi_k=(\psi_{k1},\ldots,\psi_{kJ})^T$ is a $J \times 1$ vector of cluster-specific skewness parameters; and $\boldsymbol{\epsilon}_{ki} \sim \text{N}_J(\boldsymbol0,\boldsymbol\Sigma_k)$ is a $J\times 1$ vector of correlated error terms. Thus, conditional on $t_i$ and $z_i=k$, $\boldsymbol{y}_i$ is distributed as $\text{N}_J(\mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k, \boldsymbol{\Sigma}_k)$. Marginally (after integrating over $t_i$), $\mathbf{y}_i|(z_i = k)$ is distributed $\text{MSN}_J(\boldsymbol\zeta_{ki}, \boldsymbol\alpha_k, \boldsymbol\Omega_k)$, where through back-transformation 
\begin{eqnarray}
\boldsymbol\zeta_{ki} &=& \mathbf{X}_i\boldsymbol\beta_k, \nonumber \\
\boldsymbol\alpha_k &=& \frac{1}{\sqrt{1 - \boldsymbol\psi_k^T 
\boldsymbol\Omega^{-1}_k\boldsymbol\psi_k}} \boldsymbol\omega_k \boldsymbol\Omega^{-1}_k\boldsymbol\psi_k,~~\text{and} \label{eq:back_transform}\\
\boldsymbol\Omega_k &=& \boldsymbol\Sigma_k  + \boldsymbol\psi_k \boldsymbol\psi_k^T \nonumber,
\end{eqnarray}
where $\boldsymbol\omega_k = \text{Diag}(\boldsymbol\Omega_k)^{1/2}$ is the $J \times J$ diagonal matrix containing the square root of the diagonal entries of $\boldsymbol\Omega_k$. Additional details can be found in Fr\"{u}wirth-Schnatter and Pyne (2010).

Of note, the MSN density can be expressed more compactly in terms of the matrix skew normal (MatSN) density (Chen and Gupta 2005). For cluster $k$, let $\mathbf{Y}_k$ be an ${n_k \times J}$ response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n \mathds{1}_{(z_i = k)}$ is the number of observations in cluster $k$. From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_k$ is distributed as
\begin{eqnarray}
\mathbf{Y}_{k} &\sim& \text{MatSN}_{n_k \times J}(\mathbf{M}_k,\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Omega_k)\label{eq:matsn}\\
%\text{vec}(\mathbf{Y}_k) =  (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})^T\\
\text{vec}(\mathbf{M}_k) &=& (\boldsymbol\zeta_{k1}^T,...,\boldsymbol\zeta_{kn_k}^T)^T, \nonumber
\end{eqnarray}
where $\mathbf{M}_k$ is an $n_k \times J$ location matrix with rows $\boldsymbol\zeta_{ki} = \mathbf{X}_i \boldsymbol\beta_k$ as in equation (\ref{eq:msnreg}), $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha_{kJ})^T$, $\mathbf{I}_{n_k}$ is the $n_k\times n_k$ identity matrix, and $\boldsymbol\Omega_k$ is the $J\times J$ scale matrix defined above in equation (\ref{eq:msndens}). From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_{k}$, conditional on the $n_k\times 1$ vector of random effects $\mathbf{t}_{k} = (t_1,...,t_{n_k})^T$, is jointly distributed in matrix form as
\begin{equation}
\label{eq:matnormal}
\mathbf{Y}_k | \mathbf{t}_k \sim \text{MatNorm}_{n_k \times J}(\mathbf{M}^*_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k),
\end{equation}
where $\text{MatNorm}_{n_k \times J}(\cdot)$ denotes a $n_k\times J$ matrix normal density, $\mathbf{M}^*_k$ is an $n_k \times J$ matrix such that $\text{vec}(\mathbf{M}^*_k) = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$ is an $n_kJ\times 1$ mean vector, $\mathbf{X}_k$ is an $n_kJ \times Jp$ design matrix, $\boldsymbol\beta_k$ is the $Jp\times 1$ vector of regression coefficients defined in equation (\ref{eq:msnreg}), and $\boldsymbol\Sigma_k$ is the $J\times J$ conditional covariance of $\boldsymbol\epsilon_{ik}$ given in equation (\ref{eq:msnreg}). As described in Section 3.6, the matrix representation of the MSN distribution admits convenient conjugate prior distributions for the regression parameters and scale matrices, which in turn leads to efficient Gibbs sampling for posterior inference.

\subsection{Multinomial Regression on Cluster Indicators}
\label{s:multinom}
To accommodate heterogeneity in the cluster-membership probabilities, we model $\pi_{ki}$ as a function of covariates using a multinomial logit model 
\begin{eqnarray}
\label{eq:pik}
\pi_{ki} = \Pr(z_i = k|\mathbf{w}_i) = \frac{\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k}}{\sum_{h = 1}^K \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}},~ k=1,\ldots,K,
\end{eqnarray}
where $\mathbf{w}_i$ is an $r\times 1$ vector of subject-level covariates, $\boldsymbol\delta_k$ is a $r\times 1$ vector of regression parameters associated with membership in cluster $k$. For identifiability purposes, we fix the reference category $k = K$ and set $\boldsymbol\delta_K = \mathbf{0}$. Under model (\ref{eq:pik}), $z_i|\boldsymbol\pi_i \sim \text{Multinomial}(1,\boldsymbol\pi_i)$, where $\boldsymbol\pi_i = (\pi_{1i},...,\pi_{Ki})$. As detailed below in Section \ref{s:postinf}, each cluster label $z_i$ is updated during MCMC estimation from its multinomial full conditional distribution and used in the remaining MCMC steps as the working cluster assignment for subject $i$. By allowing the cluster probabilities to vary across subjects, model (\ref{eq:mixture}) can be viewed as a \textit{mixture of experts} model, in which $\pi_{ki}$ acts as a \textit{gating function} controlling the prior probability of membership in cluster $k$, and $f(\mathbf{y}_i|\boldsymbol\theta_k)$ in equation (\ref{eq:mixture}) is the ``expert'' providing information on the within-cluster distribution of $\mathbf{y}_i$ (Bishop 2006). An appealing feature of the mixture of experts model is the ability to discern cluster membership using information contained in the covariates $\mathbf{w}_i$. That is, the gating functions make precise allocations possible even in the case when the density $f(\mathbf{y}_i | \boldsymbol\theta_k)$ is similar across clusters. In addition, modeling $\boldsymbol\pi_i$ as a function of covariates of interest allows for more robust interpretations of cluster allocations and enhances the practical interpretability of the mixture of experts model.

@@

\begin{equation}
\frac{(\text{e}^{\eta})^a}{(1 + \text{e}^{\eta})^b} = 2^{-b}\text{e}^{\kappa \eta} \int_{0}^{\infty} \text{e}^{-{w}\eta^2/2}p({w}|b,c = 0)d{w}, \label{eq:pg2}
\end{equation}
where $\kappa = a - b/2$ and $p(w|b,c = 0)$ denotes a $\text{PG}(b,0)$ density. Polson {et al.} further show that the conditional distribution $p(w|b,c)$ results from an ``exponential tilting'' of the $\text{PG}(b,0)$ density:

@@

Polson et al. use these results to show that, for the logistic model, the Bernoulli likelihood can be written as a scale-mixture of normal densities with Polya-Gamma precision terms, $w$, resulting in a normal full conditional distribution for the logistic regression parameters. To extend the P\'olya--Gamma data augmentation approach to the multinomial setting, we first introduce the binary indicators $U_{ki}$, such that $U_{ki} = \mathds{1}_{(z_i = k)}$ --- i.e., $U_{ki}$ is an indicator variable taking the value $1$ if subject $i$ belongs to cluster $k$, and $0$ otherwise. The full conditional distribution of $\boldsymbol\delta_k$ conditional on $\mathbf{U}_k = (U_{k1},...,U_{kn})^T$ and the remaining regression coefficients $\delta_h$ ($h \ne k$) is then given as  $$p(\boldsymbol\delta_k|\mathbf{U}_k,\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ki}^{U_{ki}}(1-\pi_{ki})^{1-U_{ki}},$$

@@

$$\pi_{ki} = \Pr(U_{ki} = 1) = \frac{\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}}}{1 + \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}}} = \frac{\text{e}^{\eta_{ki}}}{1 + \text{e}^{\eta_{ki}}},$$

@@

	p(\boldsymbol\delta_k|\mathbf{z},\mathbf{W}, \mathbf{O}_k) \propto p(\boldsymbol\delta_k) \text{exp} \left \{ - \frac{1}{2} (\mathbf{U}^*_k - \mathbf{W}\boldsymbol\delta_k)^T \mathbf{O}_k (\mathbf{U}^*_k - \mathbf{W}\boldsymbol\delta_k)\right \}.

@@

To accommodate intermittent missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' --- that is, as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness (i.e., observations missing not at random). Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate normal distribution given $(z_i,t_i,\mathbf{y}^{obs}_i)$:

@@

\subsubsection{Prior Specification}
We adopt a fully Bayesian inferential approach and assign prior distributions to all model parameters. Conveniently, all parameters admit conditionally conjugate priors, which greatly improves posterior computation via a data-augmented Gibbs sampler. To make use of the matrix normal representation introduced previously in Section 3, we define the $J \times (p+1)$ matrix of regression parameters $\mathbf{B}^*_k$, where $\text{vec}(\mathbf{B}^*_k) = \boldsymbol\beta^*_k = (\boldsymbol\beta_k^T,\boldsymbol\psi_k^T)^T$. We assign $\mathbf{B}^*_k|\boldsymbol\Sigma_k$ a $\text{MatNorm}(\mathbf{B}^*_{0k},\mathbf{I}_{p+1},\boldsymbol\Sigma_k)$ prior, where $\mathbf{B}^*_{0k}$ is a matrix of location parameters such that $\text{vec}(\mathbf{B}^*_{0k})$ is a vector of prior location parameters for the components of $\boldsymbol\beta^*_k$, $\mathbf{I}_{p+1}$ is the $(p+1)$-dimensional identity matrix, and $\boldsymbol\Sigma_k$ is the covariance matrix defined in equation (\ref{eq:msndens}), for which we specify an $\text{IW}(\mathbf{V}_{0k},\nu_{0k})$ prior. This leads to a matrix-normal-inverse-Wishart joint prior for $\mathbf{B}^*_k$ and $\boldsymbol\Sigma_k$, which is the conjugate joint prior for the regression parameters in the matrix normal model given in equation (\ref{eq:matsn}). This conjugate prior specification induces convenient closed-form full conditional distributions that can be easily updated within our proposed Gibbs sampler.

@@

\paragraph{\textbf{Step 2:} Update of MSN Regression Parameters} We begin the update of MSN regression parameters by first updating $t_i$, the truncated normal random effect used in the stochastic representation given in equation (\ref{eq:msnreg}). For cluster $k$, we compute $A_k = (1 + \boldsymbol\psi_k^T \boldsymbol\Sigma^{-1}_k\boldsymbol
\psi_k)^{-1}$ using current values of $\boldsymbol\psi_k$ and $\boldsymbol\Sigma_k$. Next, for $i = 1,...,n_k$ in cluster $k$, we compute $a_{ki} = A_k \boldsymbol\psi_k^T \boldsymbol\Sigma^{-1}_k(\mathbf{y}_i - \boldsymbol\zeta_{ki})$, where $\boldsymbol\zeta_{ki} = \mathbf{X}_i \boldsymbol\beta_k$. Finally, we update $t_i|(z_i = k)$ from $\text{N}_{[0,\infty)}(a_{ki}, \mathbf{A}_k)$. We repeat these updates for $k = 1,...,K$.

The remaining MSN regression parameters are updated from their full conditionals as follows. First, we form the $n_k \times (p + 1)$ matrix $\mathbf{X}^*_k$ by column-binding $\mathbf{X}_k$ and $\mathbf{t}_k$. For each cluster $k$ ($k = 1,...,K$), we update $\boldsymbol\Sigma_k$ from an $\text{IW}(\nu_k,\mathbf{V}_k)$ density, where $\nu_k = \nu_{0k} + n_k$ and\begin{eqnarray}
	\mathbf{V}_k &=&  (\mathbf{Y}_k-\mathbf{M}^*_k)^T (\mathbf{Y}_k - \mathbf{M}^*_k) + (\mathbf{B}_k^* - \mathbf{B}^*_{0k})^T (\mathbf{B}^*_k - \mathbf{B}^*_{0k}) + \mathbf{V}_{0k}, \nonumber
\end{eqnarray} where $\mathbf{M}^*_k$ is defined as in equation (\ref{eq:matnormal})
We then make use of the matrix normal representation introduced in Section 3.2 to draw $\mathbf{B}^*_k$ from a $\text{MatNorm}_{p+1,J}(\mathbb{B}_k^*,\mathbf{L}_k^{*},\boldsymbol\Sigma_k)$ density, where
\begin{eqnarray}
	\mathbb{B}^*_k &=& \mathbf{L}_k^{*} (\mathbf{X}^{*T}_k \mathbf{Y}_k + \mathbf{I}_{p+1} \mathbf{B}^*_{0k}) \nonumber \\
	\mathbf{L}^*_k &=& (\mathbf{X}^{*T}_k \mathbf{X}^*_k + \mathbf{I}_{p+1})^{-1}. \nonumber
\end{eqnarray}
For efficient sampling of the matrix normal distribution, we use the \texttt{R} package \texttt{matrixsampling} (Laurent 2018). Taking $\text{vec}(\mathbf{B}^*_k)$ yields a draw from the joint posterior of $\boldsymbol\beta_k$ and $\boldsymbol\psi_k$, and in turn $\boldsymbol\beta_k$, $\boldsymbol\alpha_k$, and $\boldsymbol\Omega_k$ via the back-transformations given in equation (\ref{eq:back_transform}).

\paragraph{\textbf{Step 3:} P\'olya--Gamma Data Augmentation for $z_i$} The sampler concludes with updates of the multinomial regression parameters $\boldsymbol\delta_k$, for $k = 1,...,K-1$, followed by updates of each latent cluster indicator $z_i$ $(i=1,\ldots,n)$ from its multinomial logit full conditional. First, we define $U_{ki} = \mathds{1}_{(z_i = k)}$ for $i = 1,...,n$ and $k = 1,...,K-1$. Next, we update ${w}_{ki}$ from a $\text{PG}(1,\eta_{ki})$ density, where $\eta_{ki} = \mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}$, and ${c}_{ki} = \log \sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$, using an efficient P\'olya-Gamma sampling algorithm implemented in the \texttt{R} package \texttt{pgdraw} (Makalic and Schmidt, 2016). Next, define $U^*_{ki} = \frac{U_{ki} - 1/2}{{w}_{ki}}$ and let $\mathbf{U}^*_k = (U^*_{k1},...,U^*_{kn})^T$. Finally, for $k = 1,...,K-1$, update $\boldsymbol\delta_k$ from a $\text{N}_r(\mathbf{d}_k,\mathbf{S}_k)$ density, where $\mathbf{S}_k = (\mathbf{S}_{k0} + \mathbf{W}^T \mathbf{O}_k \mathbf{W})^{-1}$, $\mathbf{O}_k = \text{Diag}({w}_{k1},...,{w}_{kn})$,  $\mathbf{d}_k = \mathbf{S}_k (\mathbf{S}_{k0}\mathbf{d}_{k0} + \mathbf{W}^T \mathbf{O}_k \mathbf{U}^*_k)$, and $\mathbf{W}$ is the $n \times r$ matrix of multinomial logit regression covariates such that the $i^{th}$ row of $\mathbf{W}$ is $\mathbf{w}_i$.
